---
title: "Exercise Project"
author: "Steven LaValle"
date: "April 5, 2016"
output: html_document
---
# Introduction

This is an analysis of exercise data provided by six participants in an exercise program.  The subjects were asked to perform barbell lifts correctly and incorrectly resulting in five different ways of performing the lift.  Data was collected from accelerometers on the belt, forearm, arm, and dumbell of the participants.  There are two sets of data, one for training, and one for test.  The data has a lot of missing values, not a numbers, and infinite values.  These are all set to 0.

```{r}
pml.training <- read.csv("~/Exercise/pml-training.csv")
pml.training[is.na(pml.training)] <- 0
pml.training[is.nan(unlist(pml.training))] <- 0
pml.training[is.infinite(unlist(pml.training))] <- 0
pml.testing <- read.csv("~/Exercise/pml-testing.csv")
pml.testing[is.na(pml.testing)] <- 0
pml.testing[is.nan(unlist(pml.testing))] <- 0
pml.testing[is.infinite(unlist(pml.testing))] <- 0

```

The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set.
 - Class A, exactly according to the specification
 - Class B, throwing the elbows to the front 
 - Class C, lifting the dumbbell only halfway
 - Class D, lowering the dumbbell only halfway 
 - Class E, throwing the hips to the front.

To accomplish this we have to find a combination of measurements that provide neat clusters of data. The final quiz will involve applying the model to 20 different test cases.


# Building the model from the 160 variables provided

Record Identification - these do not contribute to the analysis

X                        
user_name                
raw_timestamp_part_1    
raw_timestamp_part_2     
cvtd_timestamp           
new_window              
num_window 

Motion Data - Repeated for each sensor: belt, arm, dumbbell, forearm              

roll_<sensor>                
pitch_<sensor>              
yaw_<sensor>                 
total_accel_<sensor>         
kurtosis_roll_<sensor>      
kurtosis_picth_<sensor>      
kurtosis_yaw_<sensor>        
skewness_roll_<sensor>      
skewness_roll_<sensor>.1     
skewness_yaw_<sensor>        
max_roll_<sensor>           
max_picth_<sensor>           
max_yaw_<sensor>             
min_roll_<sensor>           
min_pitch_<sensor>           
min_yaw_<sensor>             
amplitude_roll_<sensor>     
amplitude_pitch_<sensor>     
amplitude_yaw_<sensor>       
var_total_accel_<sensor>    
avg_roll_<sensor>            
stddev_roll_<sensor>         
var_roll_<sensor>           
avg_pitch_<sensor>           
stddev_pitch_<sensor>        
var_pitch_<sensor>          
avg_yaw_<sensor>             
stddev_yaw_<sensor>          
var_yaw_<sensor>            
gyros_<sensor>_x             
gyros_<sensor>_y             
gyros_<sensor>_z            
accel_<sensor>_x             
accel_<sensor>_y             
accel_<sensor>_z            
magnet_<sensor>_x            
magnet_<sensor>_y            
magnet_<sensor>_z
           
Actual Classification - The letter A, B, C, D, E, as appropriate
classe 

The training dataset is going to consist of all of the measurements and the classe variable.  I will simply remove the first seven columns.

First I will try Principal Component Analysis to see if I can eliminate some of the variables.

```{r}
library(caret)
set.seed(12345)
inTrain <- createDataPartition(y = pml.training$classe, p = 0.70, list = FALSE)
temp.training <- pml.training[inTrain, ]
temp.training[, 160] <- as.factor(temp.training[, 160])
temp.testing <- pml.training[-inTrain, ]
temp.testing[, 160] <- as.factor(temp.testing[, 160])
training <- temp.training[, -1:-7]
testing <- temp.testing[, -1:-7]

typeColor <- ((training$classe))
# Formula used in SPAM but with factors removed.  Does not seem useful
prComp <- prcomp(training[sapply(training, function(x) !is.factor(x))])
plot(prComp$x[,1],prComp$x[,2],col=typeColor,xlab="PC1",ylab="PC2")

# Pull classe from training and try to find which variables are most significant 
pcaset <- training[, -153]
M <- abs(cor(pcaset[sapply(pcaset, function(x) !is.factor(x))]))
diag(M) <- 0
which(M > 0.8, arr.ind=T)

# With Caret
preProc <- preProcess(training[sapply(training, function(x) !is.factor(x)), -153],method="pca")
exercisePC <- predict(preProc,training[sapply(training, function(x) !is.factor(x))])

modelFit <- train(training$classe ~., method = "rpart", data=exercisePC)

confusionMatrix(training$classe,predict(modelFit,exercisePC))

test_exercisePC <- predict(preProc,testing[sapply(testing, function(x) !is.factor(x))])
confusionMatrix(testing$classe,predict(modelFit,test_exercisePC))

```



# Cross Validation

An accuracy rating of .4 is not exactly good news.

```{r}

```

# Sample Error

```{r}

```

# Rationale for the model

To develop the model, I intend to separate the individual sensors out and see if there is a direct correlation between any of the values and the classe variable.  This is suggested by the idea that movement of the waist is the primary criterion for membership in classe e - throwing the hips to the front should result in a large vector on the belt sensor and not throwing the hips forward should leave all of those values low.  The true orientation of the sensor is determined by the magnetic values, so variation of the belt magnetic values beyond a certain amount would also be a key, but there does not appear to be a way to tell what the desired orientation of the subject is.

In the final test, one record from an entire session must be evaluated to determine if the lift is being done correctly.  This means that no model that relies on knowing the previous state of the subject will be of any value.

